The Effects of weather on maize yields: New evidence from Kenya

### 2018-12-19 writing/draft3/draft16.tex
writing

To find the most suitable set of fixed effects, we adapted a stepwise selection procedure which minimized the Akaike's Information Criterion (AIC). The first step of the procedure was the estimation of the model with the complete set of our weather measures in the fixed effects and with the random intercepts. After this, we performed a search through the subsets of the fixed effects minimizing the AIC and allowing for steps in both `backward' and `forward' directions (that is removing and adding predictors). We decided not to include number of heatwave days and average maximum daily temperature although including them as fixed effects would slightly improve the AIC. The reasons for not including them were their strong correlation with average seasonal temperature (the correlation coefficient was $0.981$ in case of maximum daily temperature and $0.517$ for the number of heatwave days with the p-values smaller than $1\times10^{-6}$ in both cases), high variance inflation factor (VIF) of the maximum temperature ($5.397$) and the lack of significance of their individual t-tests (the p-values were $0.121$ for the maximum daily temperature and $0.075$ for the number of heatwave days). Furthermore, including the heatwave days and maximum daily temperature would only improve AIC from $2080.877$ to $2076.439$ (these values differ from the AIC presented in table~\ref{ARMA} in the appendix because the final estimates reported in table~\ref{ARMA} were estimated using the restricted maximum likelihood (REML) whereas the models were re-estimated using the ML method for the comparison\footnote{In general, REML is the preferred estimation method for the mixed effects models as the ML estimates of the variance component are biased, however REML criterion is not meaningful for comparison of the models with different fixed effects therefore the models need to be re-estimated by ML for comparison \cite{Zuur2009}.}). In addition, the coefficients of both heatwave days and maximum daily temperature were positive, which is opposite to what we expected. Hence, we concluded that the decrease in AIC and positive signs are likely to be results of the correlation with average temperature and so we did not include them in our preferred model.

We further applied a series of likelihood ratio tests of significance of random slopes, however no subset of random slopes improved the fit, hence, our final model only includes random intercepts.\footnote{Although adding the random slope of the coefficient of variation slightly improved the AIC from $2122.226$ to $2122.110$, we decided to not to include the coefficient of variation of temperature in the random effects. The reason for not including the random slope is that the difference in AIC is negligible and the F-test of the comparison of the two models is insignificant favouring the simpler model (p-value$=0.128$).}

To verify our results, we adapted an alternative step-down model building
approach using the Satterthwaite’s method to determine the p-values of the individual
t-tests \cite{lmerTest}. To be more specific, we started with a model which included the complete set of our weather measures in both fixed effects and random effects. In the first part of this procedure the insignificant random effects were removed one by one until only significant predictors remained. After this, the insignificant fixed effects were removed one by one until only significant predictors remained. For random effects we considered the level of significance $\alpha=0.1$ and for fixed effects we opted for $\alpha=0.05$~\cite{lmerTest}. Applying this method we obtained the same model as according to the AIC-based stepwise method described above.

@@

Based on the conditional Lagrange multiplier (LM) test developed by Baltagi \& Li (1991) and Baltagi \& Li (1995), the errors exhibited a within group autcorrelation structure in our models (the p-value of the LM test was $1.1\times10^{-13}$). To further investigate the autocorrelation structure of the errors, we estimated a number of models (with the subset of predictors chosen as described above) with ARMA($p$,$q$) error structure. In particular, we estimated all variants of our model with ARMA($p$,$q$) errors such that $p\leq2$ and $q\leq2$. Comparing the AIC criteria and using the corresponding likelihood ratio statistics, we found that the most appropriate error correlation structure was ARMA($1$,$1$). The value of the AIC criterion of the model with ARMA($1$,$1$) error structure was $2122.2$ while the AIC was $2129.2$ for the model with ARMA($1$,$0$) errors.  The p-value of the likelihood ratio test of the comparison of the model with ARMA($1$,$1$) errors and the model with ARMA($1$,$0$) errors was smaller than $1 \times10^{-4}$, hence the ARMA($1$,$1$) error structure turned out to be a better fit. The AIC criteria and the likelihood ratio tests of all the models with ARMA($p$,$q$) errors are summarised in table~\ref{ARMA} in the appendix.

>>>>

To find the most suitable set of fixed effects, we adapted a stepwise selection procedure which minimized the Akaike's Information Criterion (AIC). The first step of the procedure was the estimation of the model with the complete set of our weather measures in the fixed effects and with the random intercepts. After this, we performed a search through the subsets of the fixed effects minimizing the AIC and allowing for steps in both `backward' and `forward' directions (that is removing and adding predictors). We decided not to include number of heatwave days and average maximum daily temperature although including them as fixed effects would slightly improve the AIC. The reasons for not including them were their strong correlation with average seasonal temperature (the correlation coefficient was $0.981$ in case of maximum daily temperature and $0.517$ for the number of heatwave days with the p-values smaller than $1\times10^{-6}$ in both cases), high variance inflation factor (VIF) of the maximum temperature ($5.397$) and the lack of significance of their individual t-tests (the p-values were $0.121$ for the maximum daily temperature and $0.075$ for the number of heatwave days). Furthermore, including the heatwave days and maximum daily temperature would only lead to small change in AIC from $2080.877$ to $2076.439$ (these values differ from the AIC presented in table~\ref{ARMA} in the appendix because the final estimates reported in table~\ref{ARMA} were estimated using the restricted maximum likelihood (REML) whereas the models were re-estimated using the ML method for comparison\footnote{In general, REML is the preferred estimation method for the mixed effects models as the ML estimates of the variance component are biased, however, REML criterion is not meaningful for comparison of the models with different fixed effects. Therefore, the models had to be re-estimated by ML for comparison \cite{Zuur2009}.}). In addition, the coefficients of both heatwave days and maximum daily temperature were positive, which is opposite to what we expected. Hence, we concluded that the decrease in AIC and the positive signs are likely to be results of the correlation with average temperature and we did not include them in our preferred model.

We further applied a series of likelihood ratio tests of significance of random slopes, however no subset of random slopes improved the fit. Hence, besides the fixed effects, our final model only included random intercepts.\footnote{Although adding the random slope of the coefficient of variation slightly improved the AIC from $2122.226$ to $2122.110$, we decided to not to include the coefficient of variation of temperature in the random effects. The reason for not including the random slope is that the difference in AIC is negligible and the F-test of the comparison of the two models is insignificant favouring the simpler model (p-value$=0.128$).}

To verify our results, we adapted an alternative step-down model building
approach using the Satterthwaite’s method to determine the p-values of the individual
t-tests \cite{lmerTest}. We started with a model which included the complete set of our weather measures in both fixed effects and random effects. In the first part of the procedure, the insignificant random effects were removed one by one. In each iteration, the insignificant variable with the highest p-value was removed and this was repeated until only the significant predictors remained. After this, we applied an analogous procedure to the fixed effects until we were left only with the significant ones. As recommended in literature, we considered the level of significance $\alpha=0.1$ for random effects and $\alpha=0.05$~ for fixed effects \cite{lmerTest}. This method lead to the same model as the one obtained with our primary AIC-based method described above.

@@

According to the conditional Lagrange multiplier (LM) test developed by Baltagi \& Li (1991) and Baltagi \& Li (1995), the errors exhibited a within group autcorrelation structure in our models (the p-value of the LM test was $1.1\times10^{-13}$). To further investigate the autocorrelation structure of the errors, we estimated a number of models (with the subset of predictors chosen as described above) with an ARMA($p$,$q$) error structure. In particular, we estimated all variants of our model with ARMA($p$,$q$) errors such that $p\leq2$ and $q\leq2$. Comparing the AIC criteria and using the corresponding likelihood ratio statistics, we found that the most appropriate error correlation structure was ARMA($1$,$1$). The value of the AIC criterion of the model with the ARMA($1$,$1$) error structure was $2122.2$ while the AIC was $2129.2$ for the model with ARMA($1$,$0$) errors.  The p-value of the likelihood ratio test of the comparison of the model with ARMA($1$,$1$) errors and the model with ARMA($1$,$0$) errors was smaller than $1 \times10^{-4}$, hence the ARMA($1$,$1$) error structure turned out to be a better fit. The AIC criteria and the likelihood ratio tests of all the models with ARMA($p$,$q$) errors are summarised in table~\ref{ARMA} in the appendix.
